---
title: "FJI - 2023 Population Grid"
author: "Luis de la Rua"
date: "23-06-2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### INTRODUCTION

The present script has been created to generate a Population Grid for Fiji using 2017 Population and Housing Census (2017 PHC), which was collected using Survey Solutions CAPI, as main data input.

This is a join activity carried out by Fiji Bureau of Statistics (FBOS) and SPC's Statistics for Development Division (SDD).

The final output is a 100x100m grid that includes total population counts projected to 2023.

### POPULATION GRID SCRIPT DOCUMENTATION.
The workflow can be summarized across the following steps

#### 1. Parameters and settings
* Libraries
```{r Libraries, message=FALSE, warning=FALSE}
library(sf)
library(sp)
library(maptools)
library(raster)
library(spData)
library(rgeos)
library(dplyr)
library(rgdal)
library(magrittr)
library(terra) # using this library for spatial operations, libraries raster and rgdal will be deprecated in 2023
library(tidyterra)
```

* General Parameters and work environment
```{r WD and param}
# work directory
wd <- "C:/git/spc/popgrid_2022"
setwd(wd)

# Country code to name output data and other files
country <- 'FJI'
# Data directory
dd <- paste0("C:/Users/luisr/SPC/SDD GIS - Documents/Pacific PopGrid/UPDATE_2022/",country)

```

#### 2. Calculate the Population Growth Rate (PGR) for the current year.

```{r PGR}
# Define Population Growth Rates parameters
census_year <- 2017
current_year <- 2022
pop_2017 <- 884887  # calculated from latest census data
prjpop_2022 <-  901603   # From SDD .STAT population projections 

# Calculate PGR
popGR <- (prjpop_2022 - pop_2017)/((current_year - census_year)*pop_2017)
popGR
```

#### 3. Define Round Preserve Sum function.

This tool will help us to control the bias generated during the rounding process at pixel level. Using this function we make sure that the total population calculated from the final raster is the same as the projected population. It also makes more sens to have population values at pixel level rather than decimal numbers when we use the grid to estimate populations in certain areas.

```{r Functions}
# Define ROUND PRESERVE SUM function
round_preserve_sum <- function(x, digits = 0) {
  up <- 10 ^ digits
  x <- x * up
  y <- floor(x)
  indices <- tail(order(x-y), round(sum(x)) - sum(y))
  y[indices] <- y[indices] + 1
  y / up
}
```

#### 4. Load input datasets

* Household locations 

From from 2017 Fiji PHC they inlcude population information. We will be using Fiji Map Grid projection to avoid issues with the date line that crosses the country and split the maps when using standard projections.
```{r input datasets1, message=FALSE, warning=FALSE}
## hh locations from census
hhloc <- vect(paste0(dd,"/Population Grid Data/Fiji_HH_2017.shp"))

## We are making all the spatial analysis using Fiji Map Grid projection EPSG 3460, grabbing it from the first layer loaded
fji_crs <- crs(hhloc)

```

* EA boundaries from 2017 Fiji PHC
```{r input datasets2, message=FALSE, warning=FALSE}
## EA boundaries layer including census population and hh counts
ea <- vect(paste0(dd,"/layers/FJI_EA2017_PPGwork.gpkg"))
```

* Building Footprints (BF) extracted from Open Street Maps

Data source https://download.geofabrik.de/australia-oceania/fiji.html

This dataset is used to fill the coverage gaps that the census have in some areas of the country. We use the BF to predict where the population is more likely to be located in EAs where there are missing hh locations.
We convert the polygon layers into points to simplify the processing.
```{r input datasets3, message=FALSE, warning=FALSE}
## OSM building footprints
bf <- vect(paste0(dd,"/layers/OSM/gis_osm_buildings_a_free_1.shp"))
bf <- project(bf,fji_crs) #reproject to Fiji CRS
bf_points <- centroids(bf,inside=T) # converting the layer into centroids
```

#### 5. Fill data gaps

* Identify EAs with low coverage of household locations
```{r identify data gaps}
## extract the EAs with huge difference between census hh counts and number of HH locations to know where the data gaps can be found
eagap <- subset(ea,ea$diff > 0.5 | ea$diff < -0.5 )

## extract the bf from the original dataset that are going to be used to fill the hh locations gaps
bf_ingaps <- intersect(bf_points,eagap)
nrow(bf_ingaps)
```

* Fill the gaps with OSM building footprints

BFs do not contain population data, we assign to each point a population corresponding to the average hh size at EA level.

```{r fill data gaps}
## Merge hh locations from census with the points from the building footprints that are going to give use info on where the settlements are
hhloc_merged <- terra::union(hhloc,bf_ingaps) 
nrow(hhloc_merged)
hhloc_merged$val <- 1 #this is useful later to be able to count points in polygon

## Count number of hhlocations within each EA 
ea_simpl<-ea[,"ea2017"] # before we simplify dataset

i <- intersect (ea_simpl,hhloc_merged) # run intersection
isimp <- i[,"ea2017"] # simplify result dataframe

## Tabulate to calculate the hhcounts by EA
hhcount <- as.data.frame(isimp) %>% 
  group_by(isimp$ea2017) %>% 
  count()

## Connect hhcount table with ea layer
ea <- merge(ea,hhcount,all.x=T, by.x='ea2017', by.y='isimp$ea2017')

## Calculate Average HH size per EA (AHS)
ea$avhhsize <- ea$Total_Popu/ea$n

ea_ahs <-ea[,c("ea2017","avhhsize")]
```

#### 6. Assign AHS to all hh points so we align population counts to Census official results at EA level.

```{r Assign AHS}
## Retrieve AHS from EA and include it into hhlocations merged as an attribute 
hhloc_ahs <- intersect(i,ea_ahs)

hhloc_ahs <- hhloc_ahs[,c("ea2017","avhhsize")] # Clean unused fields from dataset

sum(hhloc_ahs$avhhsize) # check everything is adding ok
```

#### 7. Project and round populations at HH level

```{r Project and round population}
## Project the population for each of the hh locations
hhloc_ahs$pop2022 <- (hhloc_ahs$avhhsize + (hhloc_ahs$avhhsize * popGR * (as.numeric(current_year)- as.numeric(census_year))))

totpop2022 <- sum(hhloc_ahs$pop2022)

## Round to get integers
hhloc_ahs$pop2022rps <- round_preserve_sum(hhloc_ahs$pop2022)
totpop2022rps <- sum(hhloc_ahs$pop2022rps)
```


#### 8. Generate the population grid raster

```{r Generate Raster}
## Convert the Point layer into a 100x100m raster.
## Create a blank raster first
ras <- rast()
crs(ras) <- fji_crs # CRS
ext(ras) <- ext(ea)  # set extent the same as the EA fwork
res(ras) <- 100 # Resolution

rastpop2022 <- terra::rasterize(hhloc_ahs,ras,'pop2022rps',fun=sum)
# checking that raster includes same population as the original, if it 0 we are good and means that the rounding worked well 
totpop2022rps - (global(rastpop2022, fun='sum',na.rm=T)) 

## Export Raster into tif format
# writeRaster(rastpop2022,paste0(dd,"/raster/",country,"_rastpop2022_01.tif"), overwrite=T)

```

